---
title: "Performance & Scalability Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance & Scalability Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Executive Summary

Small Area Estimation often involves large administrative registries or extensive spatial networks with hundreds or thousands of domains. Traditional implementations in R often rely on interpreted loops, pure R Fisher-scoring routines, or large memory allocations for $D \times D$ dense covariance matrices.

**fastsae** re-architects core estimation algorithms in compiled C++ using `RcppArmadillo` and native `OpenMP` multi-threading, delivering:

- **Up to 364x faster** than `sae` and **12,300x faster** than `emdi` for Fay-Herriot models at $n = 1,000$.
- **Up to 80x faster** than `sae` for Spatial Fay-Herriot models at $n = 1,000$.
- **Up to 31x faster** than `sae` for Spatio-Temporal models at $n = 1,000$.
- **Massive RAM reduction**: Peak memory footprint stays under **25 MB** where existing packages require hundreds of megabytes or several gigabytes.

---

## Benchmark Results Table

The following benchmarks were conducted on simulated datasets across domain sizes ranging from $n = 30$ to $n = 1,000$ (with 5 auxiliary covariates):

| Metric | `fastsae` | `sae` (Molina & Rao) | `emdi` (Kreutzmann et al.) |
|:---|:---:|:---:|:---:|
| **Mean Time (EBLUP FH)** | **0.0015 s** | 0.291 s | 9.64 s |
| **Mean Time (Spatial FH)** | **0.165 s** | 12.60 s | 8.69 s |
| **Mean Time (Spatio-Temporal FH)** | **12.5 s** | 353.0 s | \- |
| **Peak Memory (EBLUP FH)** | **0.055 MB** | 16.3 MB | 824 MB |
| **Peak Memory (Spatial FH)** | **5.15 MB** | 408 MB | 824 MB |
| **Peak Memory (Spatio-Temporal FH)** | **0.289 MB** | 7,822 MB | \- |
| **Speedup at n = 1,000 (FH)** | **Baseline** | **~364x slower** | **~12,300x slower** |

---

## Interactive Benchmark Explorer

Use the interactive controls below to compare execution time, RAM consumption, and iterations per second across sample sizes:

```{=html}
<div class="my-4 p-3 border rounded bg-light">
  <div class="d-flex flex-wrap gap-3 mb-3">
    <div class="btn-group" role="group" aria-label="Model selection">
      <button type="button" class="btn btn-sm btn-primary active" id="btn-algo-fh" onclick="setBenchmarkAlgo('FH')">Fay-Herriot (FH)</button>
      <button type="button" class="btn btn-sm btn-outline-primary" id="btn-algo-sfh" onclick="setBenchmarkAlgo('SFH')">Spatial FH (SFH)</button>
      <button type="button" class="btn btn-sm btn-outline-primary" id="btn-algo-stfh" onclick="setBenchmarkAlgo('STFH')">Spatio-Temporal (STFH)</button>
    </div>
    <div class="btn-group" role="group" aria-label="Metric selection">
      <button type="button" class="btn btn-sm btn-success active" id="btn-metric-time" onclick="setBenchmarkMetric('time')">Execution Time</button>
      <button type="button" class="btn btn-sm btn-outline-success" id="btn-metric-mem" onclick="setBenchmarkMetric('mem')">Peak Memory</button>
      <button type="button" class="btn btn-sm btn-outline-success" id="btn-metric-ips" onclick="setBenchmarkMetric('ips')">Throughput (iter/s)</button>
    </div>
  </div>

  <div class="card mb-3 shadow-sm">
    <div class="card-header fw-bold" id="main-chart-title">
      Execution time (median, seconds) — FH · log scale
    </div>
    <div class="card-body" style="position: relative; height: 350px;">
      <canvas id="benchmark-main-chart"></canvas>
    </div>
  </div>

  <div class="card shadow-sm">
    <div class="card-header fw-bold" id="speedup-chart-title">
      Speedup Factor (fastsae vs competitors) — FH
    </div>
    <div class="card-body" style="position: relative; height: 280px;">
      <canvas id="benchmark-speedup-chart"></canvas>
    </div>
  </div>
</div>

<script src="https://cdn.jsdelivr.net/npm/chart.js"></script>
<script>
(function() {
  const RAW_DATA = [
    // FH
    { Method: "fastsae", Median: 0.000588, Mem: 0.005, IPS: 1623, n: 30, Algo: "FH" },
    { Method: "emdi", Median: 0.01088, Mem: 4.18, IPS: 91, n: 30, Algo: "FH" },
    { Method: "sae", Median: 0.00155, Mem: 0.153, IPS: 632, n: 30, Algo: "FH" },
    { Method: "fastsae", Median: 0.000511, Mem: 0.009, IPS: 1912, n: 50, Algo: "FH" },
    { Method: "emdi", Median: 0.01806, Mem: 10.47, IPS: 52, n: 50, Algo: "FH" },
    { Method: "sae", Median: 0.00195, Mem: 0.373, IPS: 494, n: 50, Algo: "FH" },
    { Method: "fastsae", Median: 0.00055, Mem: 0.016, IPS: 1761, n: 100, Algo: "FH" },
    { Method: "emdi", Median: 0.0505, Mem: 37.34, IPS: 19, n: 100, Algo: "FH" },
    { Method: "sae", Median: 0.00344, Mem: 0.918, IPS: 288, n: 100, Algo: "FH" },
    { Method: "fastsae", Median: 0.0009, Mem: 0.039, IPS: 1081, n: 250, Algo: "FH" },
    { Method: "emdi", Median: 0.857, Mem: 240.89, IPS: 1.16, n: 250, Algo: "FH" },
    { Method: "sae", Median: 0.0376, Mem: 6.35, IPS: 26, n: 250, Algo: "FH" },
    { Method: "fastsae", Median: 0.0014, Mem: 0.076, IPS: 697, n: 500, Algo: "FH" },
    { Method: "emdi", Median: 5.86, Mem: 880.31, IPS: 0.17, n: 500, Algo: "FH" },
    { Method: "sae", Median: 0.197, Mem: 18.24, IPS: 5.03, n: 500, Algo: "FH" },
    { Method: "fastsae", Median: 0.00413, Mem: 0.186, IPS: 176, n: 1000, Algo: "FH" },
    { Method: "emdi", Median: 51.02, Mem: 3773.6, IPS: 0.02, n: 1000, Algo: "FH" },
    { Method: "sae", Median: 1.504, Mem: 72.0, IPS: 0.66, n: 1000, Algo: "FH" },

    // SFH
    { Method: "fastsae", Median: 0.000735, Mem: 0.03, IPS: 1312, n: 30, Algo: "SFH" },
    { Method: "emdi", Median: 0.01071, Mem: 4.18, IPS: 93, n: 30, Algo: "SFH" },
    { Method: "sae", Median: 0.0041, Mem: 1.37, IPS: 246, n: 30, Algo: "SFH" },
    { Method: "fastsae", Median: 0.00115, Mem: 0.074, IPS: 841, n: 50, Algo: "SFH" },
    { Method: "emdi", Median: 0.018, Mem: 10.47, IPS: 44, n: 50, Algo: "SFH" },
    { Method: "sae", Median: 0.00968, Mem: 3.31, IPS: 103, n: 50, Algo: "SFH" },
    { Method: "fastsae", Median: 0.00473, Mem: 0.259, IPS: 208, n: 100, Algo: "SFH" },
    { Method: "emdi", Median: 0.053, Mem: 37.34, IPS: 17, n: 100, Algo: "SFH" },
    { Method: "sae", Median: 0.0705, Mem: 18.0, IPS: 13, n: 100, Algo: "SFH" },
    { Method: "fastsae", Median: 0.0238, Mem: 1.5, IPS: 37, n: 250, Algo: "SFH" },
    { Method: "emdi", Median: 0.844, Mem: 240.89, IPS: 1.17, n: 250, Algo: "SFH" },
    { Method: "sae", Median: 1.07, Mem: 110.39, IPS: 0.93, n: 250, Algo: "SFH" },
    { Method: "fastsae", Median: 0.142, Mem: 5.86, IPS: 6.96, n: 500, Algo: "SFH" },
    { Method: "emdi", Median: 3.14, Mem: 949.86, IPS: 0.32, n: 500, Algo: "SFH" },
    { Method: "sae", Median: 7.12, Mem: 776.4, IPS: 0.14, n: 500, Algo: "SFH" },
    { Method: "fastsae", Median: 0.832, Mem: 23.15, IPS: 1.19, n: 1000, Algo: "SFH" },
    { Method: "emdi", Median: 51.06, Mem: 3772.6, IPS: 0.02, n: 1000, Algo: "SFH" },
    { Method: "sae", Median: 67.15, Mem: 1920.0, IPS: 0.015, n: 1000, Algo: "SFH" },

    // STFH
    { Method: "fastsae", Median: 0.0367, Mem: 0.027, IPS: 25.87, n: 30, Algo: "STFH" },
    { Method: "sae", Median: 0.1186, Mem: 43.8, IPS: 7.86, n: 30, Algo: "STFH" },
    { Method: "fastsae", Median: 0.0898, Mem: 0.047, IPS: 10.92, n: 50, Algo: "STFH" },
    { Method: "sae", Median: 0.42, Mem: 107.8, IPS: 2.28, n: 50, Algo: "STFH" },
    { Method: "fastsae", Median: 2.299, Mem: 0.090, IPS: 0.434, n: 100, Algo: "STFH" },
    { Method: "sae", Median: 17.94, Mem: 2551.1, IPS: 0.056, n: 100, Algo: "STFH" },
    { Method: "fastsae", Median: 3.817, Mem: 0.225, IPS: 0.261, n: 250, Algo: "STFH" },
    { Method: "sae", Median: 58.43, Mem: 3624.7, IPS: 0.017, n: 250, Algo: "STFH" },
    { Method: "fastsae", Median: 10.36, Mem: 0.448, IPS: 0.096, n: 500, Algo: "STFH" },
    { Method: "sae", Median: 253.34, Mem: 8124.4, IPS: 0.004, n: 500, Algo: "STFH" },
    { Method: "fastsae", Median: 58.10, Mem: 0.894, IPS: 0.017, n: 1000, Algo: "STFH" },
    { Method: "sae", Median: 1786.49, Mem: 32478.3, IPS: 0.0006, n: 1000, Algo: "STFH" }
  ];

  const N_LIST = [30, 50, 100, 250, 500, 1000];
  let curAlgo = "FH";
  let curMetric = "time";
  let chartMain, chartSpeedup;

  function fmtVal(v, m) {
    if (v === null || v === undefined) return "N/A";
    if (m === "time") {
      if (v < 0.001) return (v * 1e6).toFixed(0) + " µs";
      if (v < 1) return (v * 1000).toFixed(2) + " ms";
      return v.toFixed(2) + " s";
    }
    if (m === "mem") return v.toFixed(2) + " MB";
    if (m === "ips") return v.toFixed(2) + " it/s";
    return v;
  }

  function getMetricTitle(m) {
    if (m === "time") return "Execution time (median, seconds)";
    if (m === "mem") return "Memory usage (MB)";
    if (m === "ips") return "Throughput (iterations per second)";
    return m;
  }

  window.setBenchmarkAlgo = function(algo) {
    curAlgo = algo;
    ["fh", "sfh", "stfh"].forEach(id => {
      const btn = document.getElementById("btn-algo-" + id);
      if (btn) {
        btn.classList.toggle("btn-primary", id.toUpperCase() === algo);
        btn.classList.toggle("btn-outline-primary", id.toUpperCase() !== algo);
        btn.classList.toggle("active", id.toUpperCase() === algo);
      }
    });
    updateBenchmarkCharts();
  };

  window.setBenchmarkMetric = function(m) {
    curMetric = m;
    ["time", "mem", "ips"].forEach(id => {
      const btn = document.getElementById("btn-metric-" + id);
      if (btn) {
        btn.classList.toggle("btn-success", id === m);
        btn.classList.toggle("btn-outline-success", id !== m);
        btn.classList.toggle("active", id === m);
      }
    });
    updateBenchmarkCharts();
  };

  function updateBenchmarkCharts() {
    const key = curMetric === "time" ? "Median" : curMetric === "mem" ? "Mem" : "IPS";

    const fastsaeData = N_LIST.map(n => {
      const r = RAW_DATA.find(x => x.Algo === curAlgo && x.n === n && x.Method === "fastsae");
      return r ? r[key] : null;
    });
    const saeData = N_LIST.map(n => {
      const r = RAW_DATA.find(x => x.Algo === curAlgo && x.n === n && x.Method === "sae");
      return r ? r[key] : null;
    });
    const emdiData = N_LIST.map(n => {
      const r = RAW_DATA.find(x => x.Algo === curAlgo && x.n === n && x.Method === "emdi");
      return r ? r[key] : null;
    });

    const speedupSae = N_LIST.map(n => {
      const f = RAW_DATA.find(x => x.Algo === curAlgo && x.n === n && x.Method === "fastsae");
      const s = RAW_DATA.find(x => x.Algo === curAlgo && x.n === n && x.Method === "sae");
      return f && s ? s.Median / f.Median : null;
    });
    const speedupEmdi = N_LIST.map(n => {
      const f = RAW_DATA.find(x => x.Algo === curAlgo && x.n === n && x.Method === "fastsae");
      const e = RAW_DATA.find(x => x.Algo === curAlgo && x.n === n && x.Method === "emdi");
      return f && e ? e.Median / f.Median : null;
    });

    document.getElementById("main-chart-title").textContent =
      getMetricTitle(curMetric) + " — " + curAlgo + " · log scale";
    document.getElementById("speedup-chart-title").textContent =
      "Speedup Factor (fastsae vs competitors) — " + curAlgo;

    chartMain.data.datasets[0].data = fastsaeData;
    chartMain.data.datasets[1].data = saeData;
    chartMain.data.datasets[2].data = emdiData;
    chartMain.options.scales.y.title.text = getMetricTitle(curMetric);
    chartMain.update();

    chartSpeedup.data.datasets[0].data = speedupSae;
    chartSpeedup.data.datasets[1].data = speedupEmdi;
    chartSpeedup.update();
  }

  window.addEventListener("DOMContentLoaded", function() {
    const ctx1 = document.getElementById("benchmark-main-chart");
    const ctx2 = document.getElementById("benchmark-speedup-chart");
    if (!ctx1 || !ctx2) return;

    chartMain = new Chart(ctx1.getContext("2d"), {
      type: "line",
      data: {
        labels: N_LIST.map(String),
        datasets: [
          { label: "fastsae", data: [], borderColor: "#1D6A5C", backgroundColor: "rgba(29,106,92,0.15)", tension: 0.3, fill: true, pointRadius: 5 },
          { label: "sae", data: [], borderColor: "#C77B3B", backgroundColor: "rgba(199,123,59,0.15)", tension: 0.3, fill: true, pointRadius: 5 },
          { label: "emdi", data: [], borderColor: "#7A5FA8", backgroundColor: "rgba(122,95,168,0.15)", tension: 0.3, fill: true, pointRadius: 5 }
        ]
      },
      options: {
        responsive: true,
        maintainAspectRatio: false,
        scales: {
          x: { title: { display: true, text: "Number of Domains (n)" } },
          y: { type: "logarithmic", title: { display: true, text: "" } }
        },
        plugins: {
          tooltip: {
            callbacks: {
              label: ctx => ctx.dataset.label + ": " + fmtVal(ctx.raw, curMetric)
            }
          }
        }
      }
    });

    chartSpeedup = new Chart(ctx2.getContext("2d"), {
      type: "bar",
      data: {
        labels: N_LIST.map(String),
        datasets: [
          { label: "vs sae", data: [], backgroundColor: "#C77B3B", borderRadius: 4 },
          { label: "vs emdi", data: [], backgroundColor: "#7A5FA8", borderRadius: 4 }
        ]
      },
      options: {
        responsive: true,
        maintainAspectRatio: false,
        scales: {
          x: { title: { display: true, text: "Number of Domains (n)" } },
          y: { type: "logarithmic", title: { display: true, text: "Speedup Factor (x faster)" } }
        },
        plugins: {
          tooltip: {
            callbacks: {
              label: ctx => ctx.dataset.label + ": " + (ctx.raw ? ctx.raw.toFixed(1) + "x faster" : "N/A")
            }
          }
        }
      }
    });

    updateBenchmarkCharts();
  });
})();
</script>
```

---

## Architectural Insights: Why is fastsae so Fast?

1. **Compiled C++ Linear Solvers**:
   Instead of interpreting nested loops in R, `fastsae` implements Fisher-scoring parameter search and Woodbury identity matrix inversions in Armadillo C++, directly leveraging optimized BLAS/LAPACK routines.

2. **Zero-Copy Matrix Operations**:
   Memory allocations for large intermediate structures ($V$, $V^{-1}$, $P$) are avoided or reused across Fisher iterations rather than reallocated on the heap.

3. **OpenMP Multi-Threaded Bootstrap**:
   Bootstrap resampling runs natively in parallel across CPU cores using `#pragma omp parallel for`, avoiding the serialization overhead of R worker processes (`parallel::makeCluster` / `foreach`).

4. **Woodbury Identity for Panel Data**:
   In `eblup_stfh`, inversion of the block $(DT \times DT)$ covariance matrix $V$ is reduced to operations on individual $D \times D$ and $T \times T$ blocks via Kronecker and Woodbury decomposition, transforming an $O((DT)^3)$ bottleneck into scalable $O(D^3 + T^3)$ steps.

5. **Integrated Nested Laplace Approximations (INLA)**:
   For complex hierarchical generalized and spatio-temporal models (`ebp_area`), `fastsae` utilizes INLA to compute analytical posterior margins directly from sparse Gaussian Markov Random Fields, bypassing Monte Carlo sampling entirely.

---

## Bayesian Spatio-Temporal Benchmark: `fastsae` (INLA) vs `tipsae` (Stan MCMC)

To assess Bayesian small area estimation performance, `fastsae::ebp_area()` was benchmarked against `tipsae::fit_sae()`, the state-of-the-art Stan MCMC package for spatio-temporal Beta small area models.

The test used the official `tipsae` panel dataset: **`emilia`** (38 health districts in Emilia-Romagna over 5 years, $N = 190$ domains $\times$ years) with spatial polygon contiguity matrix $W$ from **`emilia_shp`**.

| Metric | `tipsae::fit_sae` (Stan MCMC) | `fastsae::ebp_area` (INLA) | Advantage |
| :--- | :--- | :--- | :--- |
| **Model Specification** | Besag ICAR + Domain RW(1) | `spatial = "besag"`, `temporal = "rw1"`, `st_interaction = "domain-specific"` | **Exact structural equivalence** |
| **Computational Engine** | Hamiltonian Monte Carlo (NUTS Stan) | Integrated Nested Laplace Approximation | Analytical & deterministic |
| **Execution Time** | **15.8 s** (1 chain, 200 iter) / ~120 s (4 chains) | **1.93 s** (`simplified.laplace`) | **~9x to 50x+ faster** |
| **Pearson Correlation ($r$)** | Baseline | **0.9893** | **Near-identical point estimates** |
| **Mean Absolute Error (MAE)** | Baseline | **0.00304** | **Negligible numerical error** |
| **Convergence Overhead** | Requires $\hat{R} < 1.05$ checks, warmup, and tuning | None (closed-form Laplace expansions) | Instant convergence |

### Beta SAE Scalability Across Sample Sizes ($n = 30$ to $n = 1,000$)

Empirical benchmark evaluating runtime and memory consumption across expanding domain sizes:

| Number of Domains ($n$) | `fastsae` Median Time | `tipsae` Median Time | Speedup Factor | `fastsae` RAM | `tipsae` RAM | RAM Reduction |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| **$n = 30$** | **1.24 s** | 2.36 s | **1.9x faster** | 109.7 MB | 500.0 MB | **4.6x less RAM** |
| **$n = 50$** | **1.27 s** | 3.74 s | **2.9x faster** | 2.6 MB | 861.4 MB | **332x less RAM** |
| **$n = 100$** | **1.29 s** | 8.32 s | **6.4x faster** | 3.8 MB | 1,780.8 MB | **466x less RAM** |
| **$n = 250$** | **1.31 s** | 26.28 s | **20.0x faster** | 115.2 MB | 4,545.4 MB | **39.5x less RAM** |
| **$n = 500$** | **1.41 s** | 64.21 s | **45.5x faster** | 121.3 MB | 9,142.5 MB | **75.4x less RAM** |
| **$n = 1,000$** | **1.64 s** | 187.16 s | **114.2x faster** | 133.6 MB | 18,335.2 MB | **137.3x less RAM** |

```{r plot_beta_benchmark, echo = FALSE, fig.width = 9, fig.height = 4.5, out.width = "100%", message = FALSE, warning = FALSE}
library(ggplot2)
fpath <- if (file.exists("../inst/exdata/beta_benchmark.rds")) "../inst/exdata/beta_benchmark.rds" else "inst/exdata/beta_benchmark.rds"
if (file.exists(fpath)) {
  beta_df <- readRDS(fpath)
  
  p1 <- ggplot(beta_df, aes(x = n, y = `Median (s)`, color = Method, group = Method)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 3) +
    scale_color_manual(values = c("fastsae" = "#1D6A5C", "tipsae" = "#D95F02")) +
    scale_x_log10(breaks = c(30, 50, 100, 250, 500, 1000)) +
    scale_y_log10(labels = scales::label_number(suffix = " s")) +
    labs(title = "Execution Time (Log Scale)", subtitle = "Lower is faster", x = "Domains (n)", y = "Median Time (s)") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  p2 <- ggplot(beta_df, aes(x = n, y = `Memory (MB)`, color = Method, group = Method)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 3) +
    scale_color_manual(values = c("fastsae" = "#1D6A5C", "tipsae" = "#D95F02")) +
    scale_x_log10(breaks = c(30, 50, 100, 250, 500, 1000)) +
    scale_y_log10(labels = scales::label_number(suffix = " MB")) +
    labs(title = "Memory Allocation (Log Scale)", subtitle = "Lower is better", x = "Domains (n)", y = "Memory (MB)") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  if (requireNamespace("patchwork", quietly = TRUE)) {
    patchwork::wrap_plots(p1, p2, guides = "collect") & theme(legend.position = "bottom")
  }
}
```


