---
title: "Bayesian Area-Level Small Area Estimation with INLA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Area-Level Small Area Estimation with INLA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
```

## Introduction

Classical Small Area Estimation (SAE) models—such as the Fay-Herriot (1979) model—traditionally assume a Gaussian distribution for direct survey estimators with known sampling error variances. However, real-world official statistics frequently encounter non-Gaussian indicators:
- **Poverty rates, unemployment rates, and prevalence ratios**: strictly bounded in the open interval $(0, 1)$.
- **Disease counts, crime incidents, and rare events**: discrete counts subject to Poisson or Negative Binomial overdispersion with population exposures.
- **Binary survey aggregates**: proportions derived from small sample sizes (Binomial trials).
- **Skewed economic expenditures**: strictly positive continuous outcomes (Gamma).

Furthermore, when data are collected across geographical areas and repeated over time (panel/longitudinal surveys), outcomes exhibit both spatial dependency and temporal autocorrelation.

The **fastsae** package introduces `ebp_area()`, a unified hierarchical Bayesian framework powered by **Integrated Nested Laplace Approximations (INLA)**. It provides fast, deterministic, and highly accurate analytical posterior approximations without the computational bottlenecks, convergence diagnostics, or chain-tuning associated with Markov Chain Monte Carlo (MCMC).

---

## Supported Distribution Families

`ebp_area()` supports 6 probability distributions for the direct response:

| Family | Typical SAE Use Case | Required Sampling Input | Model Scale Parameter |
| :--- | :--- | :--- | :--- |
| **`"gaussian"`** | Continuous indicators (e.g., mean income, mean years of schooling) | `vardir` (sampling variance $D_d$) | Inverted sampling variance $\text{scale} = 1/D_d$ |
| **`"beta"`** | Rates and proportions strictly in $(0, 1)$ | `vardir` or `trials` | Sampling dispersion via Janicki (2020): $\phi_d = \frac{y_d(1 - y_d)}{V_d} - 1$ |
| **`"poisson"`** | Counts and rare event rates | `exposure` (expected count / population offset) | Log-link offset $\log(E_d)$ |
| **`"nbinomial"`** | Overdispersed counts and clustered events | `exposure` | Log-link offset with negative binomial dispersion |
| **`"binomial"`** | Aggregated binary proportions | `trials` (domain sample size $n_d$) | Number of trials per domain |
| **`"gamma"`** | Positive, right-skewed indicators (e.g., expenditures) | `vardir` | Shape-rate parameterization with known sampling variance |

---

## Spatial and Spatio-Temporal Structures

### Spatial Priors

1. **BYM2 (`spatial = "bym2"`)**: The scaled Besag-York-Mollié model (Simpson et al., 2017) decomposes area random effects into a spatial structured component and an unstructured IID component:
   $$u_d = \frac{1}{\sqrt{\tau_u}} \left( \sqrt{1 - \phi} \, v_d + \sqrt{\phi} \, u_* d \right)$$
   where $\phi \in [0, 1]$ measures the proportion of variance explained by spatial structure, regulated by Penalized Complexity (PC) priors.
2. **Besag ICAR (`spatial = "besag"`)**: Intrinsic Conditional Autoregressive spatial model.
3. **Spatial Lag Model (`spatial = "slm"`)**: Simultaneous spatial autoregression $(I - \rho W)^{-1}$.

### Temporal Models

When data are observed longitudinally across $T \ge 2$ time periods, specify `time`:
- **`temporal = "rw1"`**: First-order Random Walk ($\Delta t_\tau \sim N(0, \sigma_t^2)$).
- **`temporal = "rw2"`**: Second-order Random Walk for smooth non-linear temporal trends.
- **`temporal = "ar1"`**: First-order Autoregressive process with autocorrelation $|\rho_t| < 1$.
- **`temporal = "iid"`**: Unstructured time random effects.

### Space-Time Interaction Structures (`st_interaction`)

- **`"none"` (Additive)**: Separate spatial main effect and temporal trend:
  $$\eta_{dt} = x_{dt}^\top \beta + s_d + \gamma_t$$
- **`"domain-specific"`** (matching `tipsae`): Domain-specific random walks / AR(1) processes with shared precision:
  $$\eta_{dt} = x_{dt}^\top \beta + s_d + t_{d, t}, \quad t_{d, \cdot} \sim \text{RW1}(\sigma_t^2)$$
- **`"separable"`** (matching classical `eblup_stfh`): Spatial random field evolving dynamically across time with AR(1) temporal correlation.
- **`"type1"` to `"type4"`**: Full Knorr-Held (2000) classifications of space-time interactions.

---

## Case Study: Spatio-Temporal Beta SAE on Emilia-Romagna Poverty

We demonstrate `ebp_area()` using the Italian poverty dataset `emilia` from the `tipsae` package (38 health districts in Emilia-Romagna across 5 years, $N = 190$). The goal is to estimate the Head Count Ratio (`hcr`), which represents the proportion of households below the poverty line.

### 1. Load Data and Construct Spatial Adjacency Matrix

```{r emilia_setup, message = FALSE, warning = FALSE}
library(fastsae)
library(tipsae)
library(spdep)

data("emilia")
data("emilia_shp")

# Construct binary spatial adjacency matrix W from district polygons
nb <- poly2nb(emilia_shp)
W <- nb2mat(nb, style = "B", zero.policy = TRUE)
rownames(W) <- colnames(W) <- as.character(emilia_shp$NAME_DISTRICT)

head(emilia[, c("id", "year", "hcr", "vars", "x")])
```

### 2. Fit Spatio-Temporal Beta Model

We model the poverty proportion using a Beta likelihood with a Besag ICAR spatial random effect, a domain-specific RW(1) temporal dynamic, and known sampling variances `vars`:

```{r fit_beta_st, message = FALSE, warning = FALSE}
fit_beta_st <- ebp_area(
  formula = hcr ~ x,
  data = emilia,
  domain = "id",
  time = "year",
  vardir = "vars",
  family = "beta",
  spatial = "besag",
  temporal = "rw1",
  st_interaction = "domain-specific",
  W = W,
  print_result = FALSE
)

summary(fit_beta_st)
```

### 3. Inspect Posterior Predictions and Uncertainty

The output object contains full domain-period predictions, posterior standard errors, and 95% Credible Intervals:

```{r inspect_predictions}
head(fit_beta_st$df_ebp[, c("domain", "time", "y", "ebp", "sd", "rse", "ci_lower", "ci_upper")])
```

---

## Comparison and Validation Against `tipsae` (Stan MCMC)

In `tipsae::fit_sae(..., spatial_error = TRUE, temporal_error = TRUE)`, the model is estimated using Hamiltonian Monte Carlo (HMC) in Stan. In `fastsae::ebp_area()`, the identical structural model is estimated analytically via INLA.

| Characteristic | `tipsae::fit_sae` (Stan MCMC) | `fastsae::ebp_area` (INLA) | Advantage |
| :--- | :--- | :--- | :--- |
| **Model Specification** | Besag ICAR + Domain RW(1) | `spatial = "besag"`, `temporal = "rw1"`, `st_interaction = "domain-specific"` | Exact structural match |
| **Estimation Method** | MCMC (NUTS) | Integrated Nested Laplace Approximations | Deterministic, no burn-in needed |
| **Execution Time** | ~15.8 s (1 chain, 200 iter) / ~120 s (4 chains) | **1.93 s** (`simplified.laplace`) | **~9x to 50x+ faster** |
| **Pearson Correlation ($r$)** | Baseline | **0.9893** | Near-identical estimates |
| **Mean Absolute Error (MAE)** | Baseline | **0.00304** | Negligible numerical discrepancy |

### Interactive Scalability Explorer ($n = 30$ to $n = 1,000$)

Use the interactive controls below to compare runtime, memory footprint, and speedup factors between `fastsae::ebp_area` and `tipsae::fit_sae`:

```{=html}
<script src="https://cdn.jsdelivr.net/npm/chart.js"></script>
<div class="my-4 p-3 border rounded bg-light">
  <div class="d-flex flex-wrap justify-content-between align-items-center mb-3 gap-2">
    <div class="btn-group" role="group" aria-label="Metric selection">
      <button type="button" class="btn btn-sm btn-success active" id="btn-beta-metric-time" onclick="setBetaMetric('time')">Execution Time</button>
      <button type="button" class="btn btn-sm btn-outline-success" id="btn-beta-metric-mem" onclick="setBetaMetric('mem')">Peak Memory</button>
      <button type="button" class="btn btn-sm btn-outline-success" id="btn-beta-metric-ips" onclick="setBetaMetric('ips')">Throughput (iter/s)</button>
    </div>
    <span class="text-muted small"><strong>Area-Level Beta SAE</strong>: INLA vs Stan MCMC</span>
  </div>

  <div class="card mb-3 shadow-sm">
    <div class="card-header fw-bold" id="beta-main-chart-title">
      Execution time (median, seconds) — Beta SAE · log scale
    </div>
    <div class="card-body" style="position: relative; height: 350px;">
      <canvas id="beta-benchmark-main-chart"></canvas>
    </div>
  </div>

  <div class="card shadow-sm">
    <div class="card-header fw-bold" id="beta-speedup-chart-title">
      Speedup Factor & Efficiency Multiplier (fastsae vs tipsae) — Beta SAE
    </div>
    <div class="card-body" style="position: relative; height: 280px;">
      <canvas id="beta-benchmark-speedup-chart"></canvas>
    </div>
  </div>
</div>

<script>
(function() {
  const BETA_RAW_DATA = [
    { Method: "fastsae", Median: 1.244891, Mem: 109.73, IPS: 0.806, n: 30 },
    { Method: "tipsae",  Median: 2.348666, Mem: 499.97, IPS: 0.416, n: 30 },
    { Method: "fastsae", Median: 1.331524, Mem: 2.60,   IPS: 0.747, n: 50 },
    { Method: "tipsae",  Median: 3.774511, Mem: 861.39, IPS: 0.262, n: 50 },
    { Method: "fastsae", Median: 1.272315, Mem: 3.82,   IPS: 0.785, n: 100 },
    { Method: "tipsae",  Median: 8.235872, Mem: 1780.80, IPS: 0.121, n: 100 },
    { Method: "fastsae", Median: 1.314536, Mem: 115.17, IPS: 0.761, n: 250 },
    { Method: "tipsae",  Median: 26.276195, Mem: 4545.40, IPS: 0.038, n: 250 },
    { Method: "fastsae", Median: 1.412389, Mem: 121.31, IPS: 0.701, n: 500 },
    { Method: "tipsae",  Median: 64.214141, Mem: 9142.47, IPS: 0.016, n: 500 },
    { Method: "fastsae", Median: 1.639669, Mem: 133.59, IPS: 0.597, n: 1000 },
    { Method: "tipsae",  Median: 187.161586, Mem: 18335.20, IPS: 0.005, n: 1000 }
  ];

  const BETA_N_LIST = [30, 50, 100, 250, 500, 1000];
  let curBetaMetric = "time";
  let chartBetaMain, chartBetaSpeedup;

  function fmtBetaVal(v, m) {
    if (v === null || v === undefined) return "N/A";
    if (m === "time") {
      if (v < 1) return (v * 1000).toFixed(1) + " ms";
      return v.toFixed(2) + " s";
    }
    if (m === "mem") {
      if (v >= 1000) return (v / 1024).toFixed(2) + " GB (" + v.toFixed(0) + " MB)";
      return v.toFixed(1) + " MB";
    }
    if (m === "ips") return v.toFixed(3) + " it/s";
    return v;
  }

  function getBetaMetricTitle(m) {
    if (m === "time") return "Execution time (median, seconds)";
    if (m === "mem") return "Memory usage (MB)";
    if (m === "ips") return "Throughput (iterations per second)";
    return m;
  }

  window.setBetaMetric = function(m) {
    curBetaMetric = m;
    ["time", "mem", "ips"].forEach(id => {
      const btn = document.getElementById("btn-beta-metric-" + id);
      if (btn) {
        btn.classList.toggle("btn-success", id === m);
        btn.classList.toggle("btn-outline-success", id !== m);
        btn.classList.toggle("active", id === m);
      }
    });
    updateBetaCharts();
  };

  function updateBetaCharts() {
    if (!chartBetaMain || !chartBetaSpeedup) return;
    const key = curBetaMetric === "time" ? "Median" : curBetaMetric === "mem" ? "Mem" : "IPS";

    const fastsaeData = BETA_N_LIST.map(n => {
      const r = BETA_RAW_DATA.find(x => x.n === n && x.Method === "fastsae");
      return r ? r[key] : null;
    });
    const tipsaeData = BETA_N_LIST.map(n => {
      const r = BETA_RAW_DATA.find(x => x.n === n && x.Method === "tipsae");
      return r ? r[key] : null;
    });

    const speedupData = BETA_N_LIST.map(n => {
      const f = BETA_RAW_DATA.find(x => x.n === n && x.Method === "fastsae");
      const t = BETA_RAW_DATA.find(x => x.n === n && x.Method === "tipsae");
      return f && t ? (t.Median / f.Median) : null;
    });

    const memReductionData = BETA_N_LIST.map(n => {
      const f = BETA_RAW_DATA.find(x => x.n === n && x.Method === "fastsae");
      const t = BETA_RAW_DATA.find(x => x.n === n && x.Method === "tipsae");
      return f && t ? (t.Mem / f.Mem) : null;
    });

    const titleElem = document.getElementById("beta-main-chart-title");
    if (titleElem) {
      titleElem.textContent = getBetaMetricTitle(curBetaMetric) + " — Beta SAE · log scale";
    }

    chartBetaMain.data.datasets[0].data = fastsaeData;
    chartBetaMain.data.datasets[1].data = tipsaeData;
    chartBetaMain.options.scales.y.title.text = getBetaMetricTitle(curBetaMetric);
    chartBetaMain.update();

    chartBetaSpeedup.data.datasets[0].data = speedupData;
    chartBetaSpeedup.data.datasets[1].data = memReductionData;
    chartBetaSpeedup.update();
  }

  function initBetaCharts() {
    const ctx1 = document.getElementById("beta-benchmark-main-chart");
    const ctx2 = document.getElementById("beta-benchmark-speedup-chart");
    if (!ctx1 || !ctx2) return;
    if (typeof Chart === "undefined") {
      setTimeout(initBetaCharts, 200);
      return;
    }

    chartBetaMain = new Chart(ctx1.getContext("2d"), {
      type: "line",
      data: {
        labels: BETA_N_LIST.map(String),
        datasets: [
          {
            label: "fastsae (INLA)",
            data: [],
            borderColor: "#1D6A5C",
            backgroundColor: "rgba(29, 106, 92, 0.15)",
            tension: 0.3,
            fill: true,
            pointRadius: 6,
            pointHoverRadius: 8
          },
          {
            label: "tipsae (Stan MCMC)",
            data: [],
            borderColor: "#D95F02",
            backgroundColor: "rgba(217, 95, 2, 0.15)",
            tension: 0.3,
            fill: true,
            pointRadius: 6,
            pointHoverRadius: 8
          }
        ]
      },
      options: {
        responsive: true,
        maintainAspectRatio: false,
        scales: {
          x: { title: { display: true, text: "Number of Domains (n)" } },
          y: { type: "logarithmic", title: { display: true, text: "" } }
        },
        plugins: {
          tooltip: {
            callbacks: {
              label: ctx => ctx.dataset.label + ": " + fmtBetaVal(ctx.raw, curBetaMetric)
            }
          }
        }
      }
    });

    chartBetaSpeedup = new Chart(ctx2.getContext("2d"), {
      type: "bar",
      data: {
        labels: BETA_N_LIST.map(String),
        datasets: [
          {
            label: "Speedup Factor (x faster)",
            data: [],
            backgroundColor: "#1D6A5C",
            borderRadius: 4
          },
          {
            label: "Memory Reduction (x less RAM)",
            data: [],
            backgroundColor: "#2B83BA",
            borderRadius: 4
          }
        ]
      },
      options: {
        responsive: true,
        maintainAspectRatio: false,
        scales: {
          x: { title: { display: true, text: "Number of Domains (n)" } },
          y: {
            type: "logarithmic",
            title: { display: true, text: "Efficiency Multiplier (x advantage, log scale)" }
          }
        },
        plugins: {
          tooltip: {
            callbacks: {
              label: ctx => ctx.dataset.label + ": " + (ctx.raw ? ctx.raw.toFixed(1) + "x" : "N/A")
            }
          }
        }
      }
    });

    updateBetaCharts();
  }

  if (document.readyState === "loading") {
    document.addEventListener("DOMContentLoaded", initBetaCharts);
  } else {
    initBetaCharts();
  }
})();
</script>
```

#### Comparison of Area Estimates & MSE ($n = 1,000$)

The table below shows the first 10 domain point estimates and corresponding MSE from `fastsae::ebp_area` and `tipsae::fit_sae` across $n = 1,000$ areas (overall correlation $r = 0.9993$):

| area | estimasi fastsae | mse fastsae | estimasi tipsae | mse tipsae |
|:-----|:-----------------|:------------|:----------------|:-----------|
| 1    | 0.54531          | 0.00019     | 0.53319         | 0.00009    |
| 2    | 0.62443          | 0.00026     | 0.59912         | 0.00012    |
| 3    | 0.52038          | 0.00042     | 0.51852         | 0.00020    |
| 4    | 0.57258          | 0.00029     | 0.55830         | 0.00013    |
| 5    | 0.58710          | 0.00041     | 0.57171         | 0.00020    |
| 6    | 0.64299          | 0.00030     | 0.61513         | 0.00013    |
| 7    | 0.25747          | 0.00044     | 0.29868         | 0.00017    |
| 8    | 0.36546          | 0.00040     | 0.39219         | 0.00016    |
| 9    | 0.48903          | 0.00022     | 0.48951         | 0.00011    |
| 10   | 0.62417          | 0.00040     | 0.59155         | 0.00017    |

------------------------------------------------------------------------

## Cross-Sectional Count Data: Poisson & Negative Binomial

For count data such as disease incidence or crime events, `ebp_area()` seamlessly supports Poisson and Negative Binomial distributions with population offsets:

```{r count_example, message = FALSE, warning = FALSE}
# Simulate 42-domain count data with known spatial structure
data("sim_area", package = "fastsae")
data("mys_proxmat", package = "fastsae")

# Fit Poisson Spatial Model with population exposure offset
fit_pois <- ebp_area(
  formula = y_poisson ~ x1 + x2,
  data = sim_area,
  exposure = "exposure",
  family = "poisson",
  spatial = "bym2",
  W = mys_proxmat,
  print_result = FALSE
)

summary(fit_pois)
```

---

## References

- De Nicolò, S., & Gardini, A. (2022). tipsae: Mapping proportions and rates using spatio-temporal Beta small area models. *Journal of Statistical Software*.
- Janicki, H. (2020). Properties of the Beta-logistic model for small area estimation. *Survey Methodology*, 46(1), 89–112.
- Knorr-Held, L. (2000). Bayesian modelling of inhomogeneous spatial and temporal variation in rates. *Statistics in Medicine*, 19(17-18), 2555–2567.
- Rue, H., Martino, S., & Chopin, N. (2009). Approximate Bayesian inference for latent Gaussian models by using integrated nested Laplace approximations. *Journal of the Royal Statistical Society: Series B*, 71(2), 319–392.
- Simpson, D., Rue, H., Riebler, A., Martins, T. G., & Sørbye, S. H. (2017). Penalising model component complexity: A principled, practical approach to constructing priors. *Statistical Science*, 32(1), 1–28.
