---
title: "Bayesian Area-Level Small Area Estimation with INLA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Area-Level Small Area Estimation with INLA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
```

## Introduction

Classical Small Area Estimation (SAE) models—such as the Fay-Herriot (1979) model—traditionally assume a Gaussian distribution for direct survey estimators with known sampling error variances. However, real-world official statistics frequently encounter non-Gaussian indicators:
- **Poverty rates, unemployment rates, and prevalence ratios**: strictly bounded in the open interval $(0, 1)$.
- **Disease counts, crime incidents, and rare events**: discrete counts subject to Poisson or Negative Binomial overdispersion with population exposures.
- **Binary survey aggregates**: proportions derived from small sample sizes (Binomial trials).
- **Skewed economic expenditures**: strictly positive continuous outcomes (Gamma).

Furthermore, when data are collected across geographical areas and repeated over time (panel/longitudinal surveys), outcomes exhibit both spatial dependency and temporal autocorrelation.

The **fastsae** package introduces `ebp_area()`, a unified hierarchical Bayesian framework powered by **Integrated Nested Laplace Approximations (INLA)**. It provides fast, deterministic, and highly accurate analytical posterior approximations without the computational bottlenecks, convergence diagnostics, or chain-tuning associated with Markov Chain Monte Carlo (MCMC).

---

## Supported Distribution Families

`ebp_area()` supports 6 probability distributions for the direct response:

| Family | Typical SAE Use Case | Required Sampling Input | Model Scale Parameter |
| :--- | :--- | :--- | :--- |
| **`"gaussian"`** | Continuous indicators (e.g., mean income, mean years of schooling) | `vardir` (sampling variance $D_d$) | Inverted sampling variance $\text{scale} = 1/D_d$ |
| **`"beta"`** | Rates and proportions strictly in $(0, 1)$ | `vardir` or `trials` | Sampling dispersion via Janicki (2020): $\phi_d = \frac{y_d(1 - y_d)}{V_d} - 1$ |
| **`"poisson"`** | Counts and rare event rates | `exposure` (expected count / population offset) | Log-link offset $\log(E_d)$ |
| **`"nbinomial"`** | Overdispersed counts and clustered events | `exposure` | Log-link offset with negative binomial dispersion |
| **`"binomial"`** | Aggregated binary proportions | `trials` (domain sample size $n_d$) | Number of trials per domain |
| **`"gamma"`** | Positive, right-skewed indicators (e.g., expenditures) | `vardir` | Shape-rate parameterization with known sampling variance |

---

## Spatial and Spatio-Temporal Structures

### Spatial Priors

1. **BYM2 (`spatial = "bym2"`)**: The scaled Besag-York-Mollié model (Simpson et al., 2017) decomposes area random effects into a spatial structured component and an unstructured IID component:
   $$u_d = \frac{1}{\sqrt{\tau_u}} \left( \sqrt{1 - \phi} \, v_d + \sqrt{\phi} \, u_* d \right)$$
   where $\phi \in [0, 1]$ measures the proportion of variance explained by spatial structure, regulated by Penalized Complexity (PC) priors.
2. **Besag ICAR (`spatial = "besag"`)**: Intrinsic Conditional Autoregressive spatial model.
3. **Spatial Lag Model (`spatial = "slm"`)**: Simultaneous spatial autoregression $(I - \rho W)^{-1}$.

### Temporal Models

When data are observed longitudinally across $T \ge 2$ time periods, specify `time`:
- **`temporal = "rw1"`**: First-order Random Walk ($\Delta t_\tau \sim N(0, \sigma_t^2)$).
- **`temporal = "rw2"`**: Second-order Random Walk for smooth non-linear temporal trends.
- **`temporal = "ar1"`**: First-order Autoregressive process with autocorrelation $|\rho_t| < 1$.
- **`temporal = "iid"`**: Unstructured time random effects.

### Space-Time Interaction Structures (`st_interaction`)

- **`"none"` (Additive)**: Separate spatial main effect and temporal trend:
  $$\eta_{dt} = x_{dt}^\top \beta + s_d + \gamma_t$$
- **`"domain-specific"`** (matching `tipsae`): Domain-specific random walks / AR(1) processes with shared precision:
  $$\eta_{dt} = x_{dt}^\top \beta + s_d + t_{d, t}, \quad t_{d, \cdot} \sim \text{RW1}(\sigma_t^2)$$
- **`"separable"`** (matching classical `eblup_stfh`): Spatial random field evolving dynamically across time with AR(1) temporal correlation.
- **`"type1"` to `"type4"`**: Full Knorr-Held (2000) classifications of space-time interactions.

---

## Case Study: Spatio-Temporal Beta SAE on Emilia-Romagna Poverty

We demonstrate `ebp_area()` using the Italian poverty dataset `emilia` from the `tipsae` package (38 health districts in Emilia-Romagna across 5 years, $N = 190$). The goal is to estimate the Head Count Ratio (`hcr`), which represents the proportion of households below the poverty line.

### 1. Load Data and Construct Spatial Adjacency Matrix

```{r emilia_setup, message = FALSE, warning = FALSE}
library(fastsae)
library(tipsae)
library(spdep)

data("emilia")
data("emilia_shp")

# Construct binary spatial adjacency matrix W from district polygons
nb <- poly2nb(emilia_shp)
W <- nb2mat(nb, style = "B", zero.policy = TRUE)
rownames(W) <- colnames(W) <- as.character(emilia_shp$NAME_DISTRICT)

head(emilia[, c("id", "year", "hcr", "vars", "x")])
```

### 2. Fit Spatio-Temporal Beta Model

We model the poverty proportion using a Beta likelihood with a Besag ICAR spatial random effect, a domain-specific RW(1) temporal dynamic, and known sampling variances `vars`:

```{r fit_beta_st, message = FALSE, warning = FALSE}
fit_beta_st <- ebp_area(
  formula = hcr ~ x,
  data = emilia,
  domain = "id",
  time = "year",
  vardir = "vars",
  family = "beta",
  spatial = "besag",
  temporal = "rw1",
  st_interaction = "domain-specific",
  W = W,
  print_result = FALSE
)

summary(fit_beta_st)
```

### 3. Inspect Posterior Predictions and Uncertainty

The output object contains full domain-period predictions, posterior standard errors, and 95% Credible Intervals:

```{r inspect_predictions}
head(fit_beta_st$df_ebp[, c("domain", "time", "y", "ebp", "sd", "rse", "ci_lower", "ci_upper")])
```

---

## Comparison and Validation Against `tipsae` (Stan MCMC)

In `tipsae::fit_sae(..., spatial_error = TRUE, temporal_error = TRUE)`, the model is estimated using Hamiltonian Monte Carlo (HMC) in Stan. In `fastsae::ebp_area()`, the identical structural model is estimated analytically via INLA.

| Characteristic | `tipsae::fit_sae` (Stan MCMC) | `fastsae::ebp_area` (INLA) | Advantage |
| :--- | :--- | :--- | :--- |
| **Model Specification** | Besag ICAR + Domain RW(1) | `spatial = "besag"`, `temporal = "rw1"`, `st_interaction = "domain-specific"` | Exact structural match |
| **Estimation Method** | MCMC (NUTS) | Integrated Nested Laplace Approximations | Deterministic, no burn-in needed |
| **Execution Time** | ~15.8 s (1 chain, 200 iter) / ~120 s (4 chains) | **1.93 s** (`simplified.laplace`) | **~9x to 50x+ faster** |
| **Pearson Correlation ($r$)** | Baseline | **0.9893** | Near-identical estimates |
| **Mean Absolute Error (MAE)** | Baseline | **0.00304** | Negligible numerical discrepancy |

### Scalability Comparison Across Sample Sizes ($n = 30$ to $n = 1,000$)

To evaluate how computation time and memory scale with the number of domains, Area-Level Beta SAE models were benchmarked across $n = 30, 50, 100, 250, 500, 1000$:

| Number of Domains ($n$) | `fastsae` Median Time | `tipsae` Median Time | Speedup Factor | `fastsae` RAM | `tipsae` RAM | RAM Reduction |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| **$n = 30$** | **1.24 s** | 2.36 s | **1.9x faster** | 109.7 MB | 500.0 MB | **4.6x less RAM** |
| **$n = 50$** | **1.27 s** | 3.74 s | **2.9x faster** | 2.6 MB | 861.4 MB | **332x less RAM** |
| **$n = 100$** | **1.29 s** | 8.32 s | **6.4x faster** | 3.8 MB | 1,780.8 MB | **466x less RAM** |
| **$n = 250$** | **1.31 s** | 26.28 s | **20.0x faster** | 115.2 MB | 4,545.4 MB | **39.5x less RAM** |
| **$n = 500$** | **1.41 s** | 64.21 s | **45.5x faster** | 121.3 MB | 9,142.5 MB | **75.4x less RAM** |
| **$n = 1,000$** | **1.64 s** | 187.16 s | **114.2x faster** | 133.6 MB | 18,335.2 MB | **137.3x less RAM** |

```{r plot_beta_scaling, echo = FALSE, fig.width = 9, fig.height = 4.5, out.width = "100%", message = FALSE, warning = FALSE}
library(ggplot2)
fpath <- if (file.exists("../inst/exdata/beta_benchmark.rds")) "../inst/exdata/beta_benchmark.rds" else "inst/exdata/beta_benchmark.rds"
if (file.exists(fpath)) {
  beta_df <- readRDS(fpath)
  
  p1 <- ggplot(beta_df, aes(x = n, y = `Median (s)`, color = Method, group = Method)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 3) +
    scale_color_manual(values = c("fastsae" = "#1D6A5C", "tipsae" = "#D95F02")) +
    scale_x_log10(breaks = c(30, 50, 100, 250, 500, 1000)) +
    scale_y_log10(labels = scales::label_number(suffix = " s")) +
    labs(title = "Execution Time (Log Scale)", subtitle = "Lower is faster", x = "Domains (n)", y = "Median Time (s)") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  p2 <- ggplot(beta_df, aes(x = n, y = `Memory (MB)`, color = Method, group = Method)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 3) +
    scale_color_manual(values = c("fastsae" = "#1D6A5C", "tipsae" = "#D95F02")) +
    scale_x_log10(breaks = c(30, 50, 100, 250, 500, 1000)) +
    scale_y_log10(labels = scales::label_number(suffix = " MB")) +
    labs(title = "Memory Allocation (Log Scale)", subtitle = "Lower is better", x = "Domains (n)", y = "Memory (MB)") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  if (requireNamespace("patchwork", quietly = TRUE)) {
    patchwork::wrap_plots(p1, p2, guides = "collect") & theme(legend.position = "bottom")
  }
}
```

------------------------------------------------------------------------

## Cross-Sectional Count Data: Poisson & Negative Binomial

For count data such as disease incidence or crime events, `ebp_area()` seamlessly supports Poisson and Negative Binomial distributions with population offsets:

```{r count_example, message = FALSE, warning = FALSE}
# Simulate 42-domain count data with known spatial structure
data("sim_area", package = "fastsae")
data("mys_proxmat", package = "fastsae")

# Fit Poisson Spatial Model with population exposure offset
fit_pois <- ebp_area(
  formula = y_poisson ~ x1 + x2,
  data = sim_area,
  exposure = "exposure",
  family = "poisson",
  spatial = "bym2",
  W = mys_proxmat,
  print_result = FALSE
)

summary(fit_pois)
```

---

## References

- De Nicolò, S., & Gardini, A. (2022). tipsae: Mapping proportions and rates using spatio-temporal Beta small area models. *Journal of Statistical Software*.
- Janicki, H. (2020). Properties of the Beta-logistic model for small area estimation. *Survey Methodology*, 46(1), 89–112.
- Knorr-Held, L. (2000). Bayesian modelling of inhomogeneous spatial and temporal variation in rates. *Statistics in Medicine*, 19(17-18), 2555–2567.
- Rue, H., Martino, S., & Chopin, N. (2009). Approximate Bayesian inference for latent Gaussian models by using integrated nested Laplace approximations. *Journal of the Royal Statistical Society: Series B*, 71(2), 319–392.
- Simpson, D., Rue, H., Riebler, A., Martins, T. G., & Sørbye, S. H. (2017). Penalising model component complexity: A principled, practical approach to constructing priors. *Statistical Science*, 32(1), 1–28.
