---
title: "Model Diagnostics and Residual Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Diagnostics and Residual Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
```

## Introduction

In Small Area Estimation (SAE), model-based estimators borrow strength across domains and time to improve estimation precision over direct survey estimators. However, introducing model assumptions carries the risk of model misspecification, bias, and inappropriate shrinkage.

To address these challenges, **fastsae** provides a unified diagnostic framework centered on the `diagnose()` function and its dedicated S3 `autoplot()` visualization method. It evaluates both **frequentist C++ EBLUP** models (`eblup_fh`, `eblup_sfh`, `eblup_stfh`, `eblup_bhf`) and **Bayesian INLA EBP** models (`ebp_area`).

---

## Core Diagnostic Criteria

The diagnostic workflow evaluates four fundamental dimensions of SAE model quality:

1. **Precision Gain & RSE Reduction**:
   Quantifies the improvement in Relative Standard Error ($\text{RSE} = \frac{\sqrt{\text{MSE}}}{\hat{\theta}} \times 100\%$) compared to direct survey estimates. Evaluates whether areas satisfy the official statistics reliability threshold (typically $\text{RSE} < 25\%$ or $20\%$).
2. **External Calibration & Bias Diagnostic (Brown et al., 2001)**:
   Tests whether model predictions are statistically unbiased predictors of direct estimates using the regression:
   $$y_d = \alpha + \beta \hat{\theta}_d + e_d$$
   Under unbiasedness, the joint hypothesis $H_0: \alpha = 0, \beta = 1$ should not be rejected.
3. **Goodness-of-Fit Statistic**:
   Computes the Chi-square diagnostic statistic $W$:
   $$W = \sum_{d=1}^D \frac{(y_d - \hat{\theta}_d)^2}{V_d + \text{MSE}(\hat{\theta}_d)}$$
   which approximately follows a $\chi^2(D)$ distribution under the null hypothesis of proper model specification.
4. **Residual Spatial Autocorrelation (Moran's I)**:
   Verifies whether spatial autocorrelation has been fully captured by the model random effects. Significant residual spatial clustering indicates that unmodeled spatial patterns remain.

---

## Practical Example: Diagnosing a Fay-Herriot Model

We demonstrate the diagnostic workflow using the built-in `mys` dataset (Mean Years of Schooling in 42 regencies):

### 1. Fit Model and Run `diagnose()`

```{r diag_fit, message = FALSE, warning = FALSE}
library(fastsae)
library(ggplot2)

data("mys")
data("mys_proxmat")

# Fit Area-Level Fay-Herriot model
fit_fh <- eblup_fh(
  formula = y ~ x1 + x2 + x3,
  vardir = ~vardir,
  data = na.omit(mys),
  method = "REML",
  print_result = FALSE
)

# Run comprehensive diagnostic checks
diag_fh <- diagnose(fit_fh, W = mys_proxmat[!is.na(mys$y), !is.na(mys$y)])
print(diag_fh)
```

### 2. Inspect Diagnostic Components

The resulting `fastsae_diagnose` object provides structured access to each diagnostic domain:

```{r diag_precision}
# Efficiency gains
diag_fh$precision

# Calibration test results
diag_fh$brown_test

# Moran's I spatial test on residuals
diag_fh$spatial_test
```

---

## Visual Diagnostics with `autoplot()`

`fastsae` implements dedicated ggplot2 visualization types for `fastsae_diagnose` objects:

### 1. Calibration Plot (`type = "calibration"`)

Plots direct survey estimates against model predictions along with the 1:1 identity line and the fitted calibration regression:

```{r plot_calib}
autoplot(diag_fh, type = "calibration")
```

### 2. Precision Comparison (`type = "rse"`)

Visualizes the domain-by-domain reduction in Relative Standard Error:

```{r plot_rse}
autoplot(diag_fh, type = "rse")
```

### 3. Residual Diagnostics (`type = "residuals"` & `type = "qq"`)

Inspect standardized residuals against fitted values to detect heteroscedasticity, non-linearities, or outliers:

```{r plot_res}
autoplot(diag_fh, type = "residuals")
autoplot(diag_fh, type = "qq")
```

### 4. Combined Diagnostic Dashboard (`type = "all"`)

Generates a unified diagnostic view summarizing calibration and precision gains:

```{r plot_all}
autoplot(diag_fh, type = "all")
```

---

## Diagnosing Bayesian Spatio-Temporal Models

The exact same `diagnose()` function applies directly to Bayesian models fitted with `ebp_area()`:

```{r diag_ebp, message = FALSE, warning = FALSE}
data("sim_area", package = "fastsae")

# Fit Spatial EBP model with BYM2 prior
fit_ebp <- ebp_area(
  formula = y_gaussian ~ x1 + x2,
  data = sim_area,
  vardir = "vardir",
  spatial = "bym2",
  W = mys_proxmat,
  print_result = FALSE
)

# Diagnose Bayesian model
diag_ebp <- diagnose(fit_ebp)
diag_ebp$precision
diag_ebp$status
```

---

## References

- Brown, G., Chambers, R., Heady, P., & Heasman, D. (2001). Evaluation criteria for small area estimation methods. *Statistics in Transition*, 5(2), 185–200.
- Fay, R. E., & Herriot, R. A. (1979). Estimates of income for small places: An application of James-Stein procedures to Census data. *Journal of the American Statistical Association*, 74(366), 269–277.
- Rao, J. N. K., & Molina, I. (2015). *Small Area Estimation* (2nd ed.). John Wiley & Sons.
