---
title: "Spatial and Spatio-Temporal Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial and Spatio-Temporal Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

When small areas are geographic units (such as counties, regencies, or districts), spatial correlation frequently occurs: neighboring areas tend to have more similar outcomes than distant ones. Furthermore, when surveys are repeated over multiple years, temporal correlation between time periods is also present.

**fastsae** provides specialized, high-performance implementations for both:
1. **Spatial Fay-Herriot (`eblup_sfh`)**: Simultaneous Autoregressive SAR(1) random effects.
2. **Spatio-Temporal Fay-Herriot (`eblup_stfh`)**: Combined spatial SAR(1) and temporal AR(1) random effects.

---

## 1. Spatial Fay-Herriot Model (`eblup_sfh`)

### Model Formulation

The Spatial Fay-Herriot model (Pratesi & Salvati, 2008) models area random effects using a Simultaneous Autoregressive (SAR(1)) process:

$$y = X\beta + u + e$$

$$u = \rho_1 W u + \epsilon_1 \implies u = (I - \rho_1 W)^{-1} \epsilon_1$$

where:
- $W$ is a known row-standardized spatial proximity matrix ($D \times D$).
- $\rho_1 \in (-1, 1)$ is the spatial autoregressive parameter.
- $\epsilon_1 \sim N(0, \sigma_{u1}^2 I_D)$ is independent innovation noise.
- $e \sim N(0, V_e)$ is sampling error with diagonal matrix $V_e = \text{diag}(D_1, \dots, D_D)$.

### Fitting the Model

```{r sfh_fit}
library(fastsae)
data("mys")
data("mys_proxmat")

# Fit Spatial Fay-Herriot model with REML
fit_sfh <- eblup_sfh(
  formula = y ~ x1 + x2 + x3,
  vardir  = ~vardir,
  data    = mys,
  W       = mys_proxmat,
  method  = "REML",
  print_result = FALSE
)

summary(fit_sfh)
```

### Unsampled Domains and Spatial Kriging

A major advantage of `eblup_sfh` in **fastsae** is its automatic handling of unsampled domains (domains with `y = NA`). Instead of throwing an error or requiring manual subsetting, `eblup_sfh` automatically performs **full-spatial kriging**:

```{r unsampled}
# Count sampled vs unsampled domains
table(is.na(mys$y))

# Inspect estimates for unsampled domains
head(fit_sfh$df_eblup[is.na(mys$y), ])
```

For unsampled areas, prediction borrows strength from both the regression synthetic component $x_d^\top \hat{\beta}$ and spatial proximity to neighboring sampled areas through $(I - \hat{\rho}_1 W)^{-1}$.

### Multi-Threaded Bootstrap MSE

In addition to analytical Prasad-Rao style MSE, `eblup_sfh` supports multi-threaded OpenMP bootstrap MSE estimation:

- **Parametric Bootstrap (`mse_method = "pbmse"`)**
- **Non-Parametric Bootstrap (`mse_method = "npbmse"`)**

```{r sfh_pbmse}
fit_sfh_pb <- eblup_sfh(
  formula    = y ~ x1 + x2 + x3,
  vardir     = ~vardir,
  data       = mys,
  W          = mys_proxmat,
  mse_method = "pbmse",
  B          = 50,
  n_threads  = 2,
  seed       = 123,
  print_result = FALSE
)

head(fit_sfh_pb$df_eblup[, c("domain", "y", "eblup", "mse", "mse_pb", "mse_pbbc")])
```

---

## 2. Spatio-Temporal Fay-Herriot Model (`eblup_stfh`)

### Model Formulation

The Spatio-Temporal model (Marhuenda, Molina, & Morales, 2013) combines spatial correlation and temporal autoregression across a balanced panel of $D$ areas observed over $T$ time periods:

$$y_{dt} = x_{dt}^\top \beta + u_{1d} + u_{2dt} + e_{dt}$$

where:
- $u_1 = (u_{11}, \dots, u_{1D})^\top$ captures static spatial effects: $u_1 = \rho_1 W u_1 + \epsilon_1$.
- $u_{2d} = (u_{2d1}, \dots, u_{2dT})^\top$ captures dynamic temporal effects for area $d$: $u_{2dt} = \rho_2 u_{2d,t-1} + \epsilon_{2dt}$.
- $e_{dt} \sim \text{ind. } N(0, D_{dt})$ are sampling errors.

### Fitting the Spatio-Temporal Model

```{r stfh_fit}
data("mys_panel")

# Prepare balanced panel without missing values
panel_data <- mys_panel[!is.na(mys_panel$y) & mys_panel$year >= 2024, ]
W_sub <- mys_proxmat[-c(21, 25), -c(21, 25)]

# Fit Spatio-Temporal model with bootstrap MSE
fit_stfh <- eblup_stfh(
  formula     = y ~ x1 + x2 + x3,
  data        = panel_data,
  vardir      = ~vardir,
  domain      = ~area,
  time        = ~year,
  W           = W_sub,
  model       = "ST",
  compute_mse = TRUE,
  B           = 25,
  seed        = 42,
  print_result = FALSE
)

summary(fit_stfh)
```

The estimated variance and correlation components include:
- `sigma21`: Spatial random effect variance ($\sigma_{u1}^2$).
- `rho1`: Spatial autocorrelation ($\rho_1$).
- `sigma22`: Temporal random effect variance ($\sigma_{u2}^2$).
- `rho2`: Temporal AR(1) autoregression coefficient ($\rho_2$).

---

## 3. Comparing Models with `autoplot`

You can directly compare EBLUP estimates and uncertainty across different models using `autoplot()`:

```{r compare_models}
# Compare Fay-Herriot vs Spatial Fay-Herriot
fit_fh <- eblup_fh(y ~ x1 + x2 + x3, vardir = ~vardir, data = na.omit(mys), print_result = FALSE)
W_clean <- mys_proxmat[!is.na(mys$y), !is.na(mys$y)]
fit_sfh_clean <- eblup_sfh(y ~ x1 + x2 + x3, vardir = ~vardir, data = na.omit(mys), W = W_clean, print_result = FALSE)

autoplot(list("Standard FH" = fit_fh, "Spatial FH" = fit_sfh_clean), type = "comparison")
```

## References

- Marhuenda, Y., Molina, I., & Morales, D. (2013). Small area estimation with spatio-temporal Fay-Herriot models. *Computational Statistics & Data Analysis*, 58, 308–325.
- Pratesi, M., & Salvati, N. (2008). Small area estimation for spatially correlated data: A Fay-Herriot with the spatial linear spline model. *Journal of Applied Statistics*, 35(7), 781–794.
