---
title: "Unit-Level Estimation with Battese-Harter-Fuller"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unit-Level Estimation with Battese-Harter-Fuller}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

Unlike area-level models that aggregate data prior to modeling, **unit-level models** operate directly on individual survey unit records (e.g., households, farms, or persons) while linking them to population auxiliary aggregates (e.g., census means or satellite imagery).

The **fastsae** package implements the nested error regression model of Battese, Harter, and Fuller (1988) via `eblup_bhf()`, providing fast estimation and parallel parametric bootstrap MSE.

---

## Model Formulation

For individual unit $j$ ($j = 1, \dots, n_d$) in small area $d$ ($d = 1, \dots, D$):

$$y_{dj} = x_{dj}^\top \beta + u_d + e_{dj}$$

where:
- $u_d \sim \text{i.i.d. } N(0, \sigma_u^2)$ is the area-specific random effect.
- $e_{dj} \sim \text{i.i.d. } N(0, \sigma_e^2)$ is the unit-level error variance.
- $u_d$ and $e_{dj}$ are mutually independent.

The small area population mean $\bar{Y}_d$ is estimated by:

$$\hat{\bar{Y}}_d^{\text{EBLUP}} = \bar{X}_d^\top \hat{\beta} + \gamma_d (\bar{y}_d - \bar{x}_d^\top \hat{\beta})$$

where:
- $\bar{X}_d$ is the known population mean vector of auxiliary variables for domain $d$.
- $\bar{y}_d$ and $\bar{x}_d$ are the sample means for domain $d$.
- $\gamma_d = \frac{\sigma_u^2}{\sigma_u^2 + \sigma_e^2 / n_d}$ is the shrinkage ratio.

---

## Step-by-Step Example

### 1. Data Preparation

We use the classic `cornsoybean` dataset, reporting corn crop hectares per segment in 12 Iowa counties:

```{r data_prep}
library(fastsae)
data("cornsoybean")
data("cornsoybeanmeans")

# Align column names for population auxiliary means
df_pop <- cornsoybeanmeans
names(df_pop)[names(df_pop) == "MeanCornPixPerSeg"] <- "CornPix"
names(df_pop)[names(df_pop) == "MeanSoyBeansPixPerSeg"] <- "SoyBeansPix"
names(df_pop)[names(df_pop) == "CountyIndex"] <- "County"

head(cornsoybean)
head(df_pop)
```

### 2. Fit BHF Model with Bootstrap MSE

We fit the model using `eblup_bhf()`. To estimate domain-level Mean Squared Error (MSE), set `mse = TRUE`:

```{r fit_bhf}
fit_bhf <- eblup_bhf(
  formula     = CornHec ~ CornPix + SoyBeansPix,
  unit_data   = cornsoybean,
  Xpop        = df_pop,
  domain_var  = "County",
  popsize_var = "PopnSegments",
  method      = "REML",
  mse         = TRUE,
  B           = 50,
  seed        = 123,
  print_result = FALSE
)

summary(fit_bhf)
```

### 3. Inspect Domain Estimates

The resulting `df_eblup` data frame contains the domain estimates along with sample size, MSE, and Relative Standard Error (RSE):

```{r eblup_table}
head(fit_bhf$df_eblup)
```

### 4. Visualizing Results with `autoplot`

For unit-level models, domain uncertainty and model comparisons can be displayed with `autoplot()`:

```{r plot_bhf}
# Plot MSE across counties
autoplot(fit_bhf, type = "mse")

# Plot EBLUP estimates with confidence intervals
autoplot(fit_bhf, type = "comparison")
```

## References

- Battese, G. E., Harter, R. M., & Fuller, W. A. (1988). An error-components model for prediction of county crop areas using survey and satellite data. *Journal of the American Statistical Association*, 83(401), 28–36.
- Rao, J. N. K., & Molina, I. (2015). *Small Area Estimation* (2nd ed.). John Wiley & Sons.
